Using SIMD and SIMT vectorization to evaluate sparse chemical kinetic Jacobian matrices and thermochemical source terms

### 2018-06-04 paper.tex
edits through page 7

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform, \texttt{pyJac}, for single-instruction, multiple-data (SIMD) and single-instruction, multiple thread (SIMT) vectorized \revise{constant-pressure\slash volume} thermo-chemical source-term and sparse\slash dense chemical kinetic Jacobian evaluation has been developed.% Can you introduce this a bit more? good to start with at least one sentence of introduction. also, this is passive and quite long\slash complex
Selected chemical kinetic models covering the range of sizes typically used in reactive-flow simulations were used for demonstration.
A new formulation of the chemical kinetic governing equations was derived and validated, \revise{resulting in} Jacobian sparsities of \SIrange{28.6}{92.0}{$\percent$} for the tested models.
Speedups of \SIrange{3.40}{4.08}{$\times$} were found for shallow-vectorized OpenCL source-rate evaluation \revise{compared with} a parallel OpenMP code on an \avx/ central processing unit (CPU), increasing to \SIrange{6.63}{9.44}{$\times$} and \SIrange{3.03}{4.23}{$\times$} for sparse and dense chemical kinetic Jacobian evaluation, respectively.
In addition, the performance of a Nvidia \gpunew/ graphics processing unit (GPU) was compared to a Nvidia \gpuold/ GPU finding speedups of \SIrange{1.40}{1.88}{$\times$}, \SIrange{1.10}{1.59}{$\times$}, and \SIrange{1.36}{3.0}{$\times$} for source-rate, sparse, and dense Jacobian evaluation, respectively. %is this sentence necessary in the abstract?
Furthermore, the effect of data-ordering was investigated and \revise{a} storage pattern specifically formulated for vectorized evaluation was proposed; as well, the effect of the \revise{constant pressure\slash volume} assumptions and varying vector widths were studied on source-term evaluation performance.
Speedups over a first-order finite-difference approach reached \SIrange{17.60}{45.13}{$\times$} and \SIrange{55.11}{245.63}{$\times$} for sparse\slash dense evaluation on the CPU \slash GPU respectively,%not sure what numbers refer to- sparse, dense, CPU, GPU?
while dense Jacobian evaluation was up to \SI{19.56}{$\times$} and \SI{2.84}{$\times$} times faster than a previous version of \texttt{pyJac} on a CPU and GPU, respectively.
Finally, future directions for vectorized chemical kinetic evaluation and sparse linear-algebra techniques were discussed.
\end{abstract}

@@

A host of techniques have been developed to lessen the computational demand of chemical kinetic calculations while maintaining fidelity, \revise{falling} broadly into \revise{three} categories: removal of unimportant species and reactions~\cite{Lu:2006bb,Pepiot-Desjardins:2008,Hiremath:2010jw,Niemeyer:2010bt,Curtis:2015},
lumping of species with similar thermo-chemical properties~\cite{Lu:2007,Ahmed:2007fa,Pepiot:2008kq},
and time-scale methods that reduce numerical stiffness~\cite{Maas:1992ws,Lam:1994ws,Lu:2001ve,Gou:2010}.
\revise{We refer interested readers} to the recent review by Tur\'anyi and Tomlin~\cite{turanyi2016analysis} for a comprehensive overview.

@@

Recognizing the need to accelerate chemical-kinetic Jacobian evaluation and factorization, a number of recent works have been published on \revise{constructing analytical Jacobian matrices}; although as will be discussed at the end of this section, \revise{here we offer} several key improvements over past efforts.
\add{Schwer et al.~\cite{SCHWER2002270} were among the first to recognize the critical importance of a sparse analytical Jacobian to accelerate chemical kinetic simulations.}
The \texttt{TChem} software package~\cite{Safta:2011vn} was one of the first codes developed providing analytical Jacobian evaluation, but has several limitations including: incompatibilities with modern reaction types---i.e., pressure dependent Arrhenius (or P-Log) and Chebyshev reactions---and lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}.
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility; although the chosen isothermal assumption is not typical in most combustion simulations, a significant increase in Jacobian sparsity was exhibited.
Additionally, in the same work Bisetti provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without need for storage of the full dense Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at expense of strict correctness), and enable tabulation of temperature dependent properties; an \SI{80}{\percent} speedup over a finite-difference-based Jacobian was reported when used in a multidimensional reactive-flow simulation.
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and found speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species) using a Tesla C2075 GPU.
\add{Bauer et al.~\cite{Bauer:2014} used warp-specialization to improve GPU-vectorization over a standard data-parallel vectorization approach; speedups up to \SIrange{2.81}{3.75}{$\times$}, \SIrange{1.91}{2.58}{$\times$}, and \SIrange{1.4}{1.5}{$\times$} were achieved on a Telsa K20c GPU\todo{keep hardware spec?} for evaluation of viscosity, species diffusion and chemical source terms, respectively}.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, supporting parallel execution on the CPU and SIMT execution on the GPU; a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian was found on the CPU.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but it was not validated outside the context of use with an implicit-integration technique; in addition, since the Jacobian was based on an over-constrained system~\cite{HANSEN2018257}, the effect on strict conservation of mass\slash energy was not studied.
\add{Hansen and Sutherland~\cite{HANSEN2018257} explored the choice of thermo-chemical state-vectors and their resulting effect on consistency and errors in conserved properties such as mass and energy; further, the effect on implicit\slash linearly-implicit integration techniques and analysis techniques such as chemical explosive mode analysis~\cite{lu_yoo_chen_law_2010} were characterized.}

\todo[inline]{I don't know that I actually want to remove the processor information alone here -- it doesn't save us much space.  It would be more valuable to re-write this entirely, but I'm not sure that's necessary either}
A number of recent works have investigated the use of high-performance SIMT-devices to accelerate reactive-flow and chemical kinetics simulations.
Spafford et al.~\cite{Spafford:2010aa} coupled GPU-based chemical source-term evaluation with an explicit direct numerical simulation code, achieving an order of magnitude speedup \delete{on a Tesla C1060} compared to a serial implementation \add{CPU}\delete{on a AMD-Opteron processor}.
Shi et al.~\cite{Shi:2011aa} combined GPU-based chemical kinetic source-term evaluation and Jacobian factorization with two implicit CPU solvers, achieving an order of magnitude speedup for homogeneous reactor simulations of large chemical models\delete{on a Tesla 2050 GPU} over a serial\delete{Intel i7 930} CPU implementation.
Niemeyer et al.~\cite{Niemeyer:2011aa} implemented an explicit solver for non-stiff chemistry on a Tesla C2075 GPU, achieving a speedup of nearly two orders of magnitude over a sequential \add{CPU} code\delete{on an Intel Xeon X5650 CPU}.
Shi et al.~\cite{Shi:2012aa} proposed a strategy for chemical-kinetic integration in three-dimensional reactive-flow simulations, a traditional CPU-based implicit integrator handled the stiffest computational cells while a stabilized-explicit GPU-solver\delete{on a Tesla C2050 GPU} solved the less stiff cells; this hybrid solution technique was \SIrange{11}{46}{$\times$} faster than the implicit CPU solver alone for simulation of a premixed diesel engine.
Le et al.~\cite{Le2013596} found a \SIrange{30}{50}{$\times$} speedup for a GPU-based shock-capturing reactive-flow code\delete{ on a Tesla C2070} as compared to a sequential \add{CPU} version of the same\delete{on an Intel Xeon X5650 CPU}.
Stone and Davis~\cite{Stone:2013aa} investigated a GPU-version of a common implicit integrator (VODE~\cite{Brown:1989vl}), finding an order of magnitude speedup over a serial CPU implementation.
Niemeyer and Sung~\cite{Niemeyer:2014aa} developed a GPU-based stabilized explicit integrator for use with moderately-stiff chemical kinetics, achieving an order of magnitude speedup\delete{on a Tesla C2075 GPU} over a multithreaded VODE solver on a six-core\delete{Intel X5650} CPU.
Sewerin and Rigopoulos~\cite{Sewerin20151375} studied a fifth-order implicit Runge-Kutta solver on both consumer-grade and high-end GPUs\slash CPUs; the high-end GPU solver was at best \SI{1.8}{$\times$} slower than the high-end CPU version running on \num{16} cores.
Yonkee and Sutherland~\cite{Yonkee2016} implemented GPU-accelerated evaluations of thermodynamic parameters, multicomponent transport properties and species production rates for solution of a partial differential equations (PDEs)\delete{on an Intel Xeon E5-2680} \add{both the CPU} and\delete{a Tesla K20} GPU.
In evaluation of the thermo-chemical properties, speedups between \SIrange{8}{13}{$\times$} were found (over a serial version of the same code) while running on 16 CPU cores, and \SIrange{20}{40}{$\times$} on the GPU; speedups of \textasciitilde\SI{9}{$\times$} and \textasciitilde\SI{25}{$\times$} were found in solution of a partially premixed methanol flame PDE on 16 CPU cores and the GPU, respectively.
Curtis et al.~\cite{CurtisGPU:2017} implemented a fifth-order implicit Runge--Kutta method~\cite{wanner1991solving}, as well as two fourth-order exponential integration techniques~\cite{Hochbruck:1998,Hockbruck:2009} paired with an analytical Jacobian code~\cite{Niemeyer:2016aa}\delete{on a Tesla C2075} \add{on the GPU and CPU}; the \add{GPU-based} implicit Runge--Kutta method was roughly equivalent to a standard implicit integrator~\cite{Hindmarsh:2005} running on \numrange{12}{38}\delete{Intel Xeon E5-4640 v2} CPU cores for two relatively small chemical models with an integration time step of \SI{e-6}{$\sec$}.

@@

When storing arrays for a chemical kinetic model, the data-storage layout and vectorization patterns are critical to achieving high-performance code.
\Cref{F:mem} depicts an example data array with $K$ rows and $N$ columns, and index ($i$, $j$) corresponding to the $i$th row and $j$th column.
For example, the concentration of species $j$ for the $i$th thermo-chemical state would be stored in $[C]_{i, j}$ with $1 \le i \le N_{state}$ (the number of thermo-chemical states considered for evaluation), and $1 \le j \le \ns/$ (the number of species in the model).
The stored concentrations would then have $K = N_{\text{state}}$ rows, and $N = N_{\text{sp}}$ columns.

In the ``C'' (C row-major) format, the concentrations of all species for a single thermo-chemical condition $i$ would be stored sequentially in memory, i.e., with  $[C]_{1, 1}$ in index \num{1}, $[C]_{1, 2}$ in index \num{2}, etc., as shown in \cref{F:row_major}.
Conversely in the ``F'' (Fortran column-major) format, the concentrations of a single species $j$ over all thermo-chemical states are adjacent in memory, corresponding to storing $[C]_{1, 1}$ in index \num{1}, $[C]_{2, 1}$ in index \num{2} and so on, as shown in \cref{F:column_major}.
This ordering---along with the device (CPU, GPU, etc.\@) and vectorization pattern in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD\slash SIMT vectorization (also referred to as ``per-thread'' in previous works~\cite{Stone:2013aa} using GPUs), each SIMD-lane (or SIMT-thread) in a compute-unit evaluates the thermo-chemical source-terms\slash chemical kinetic Jacobian for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes\slash SIMT-thread accessing $[C]_{1, j}\ldots[C]_{N_v, j}$---the concentration of species $j$ for states $1, 2,\ldots N_v$, where $N_v$ is the SIMD vector-width or the number of threads in a SIMT warp---will load sequential locations in memory.
The first $(j+1)$th species concentration, $[C]_{1, j+1}$, will be $N_{\text{state}}$ memory locations away; this increases the likelihood of cache-misses on the CPU~\cite{gray2000rules}, but conversely is well suited to coalesced memory accesses on the GPU~\cite{Nvidia:2018}.

In a \textit{deep}-SIMD\slash SIMT vectorization (also referred to as ``per-block'' in previous GPU works (e.g.,~\cite{Stone:2013aa,CurtisGPU:2017}), a compute-unit utilizes its SIMD-lanes\slash SIMT-threads cooperatively to evaluate the thermo-chemical source-terms for a single thermo-chemical state; thus SIMD-lanes loading $[C]_{1, j} \ldots [C]_{1, j + N_v}$---the species concentrations for state 1, species $j \ldots j + N_v$---will access sequential memory locations if the data is stored in ``C''-order.
Further, in ``C''-ordering the maximum distance between any two species concentrations within the same thermo-chemical state is $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU, but may lead to uncoalesced memory-accesses on the GPU.
In addition, a deep vectorization requires both synchronization between SIMD-lanes\slash SIMT-threads via memory-fences\slash barriers (an expensive operation), and may result in SIMD-waste\slash SIMT thread-divergence, caused by different SIMD-lanes\slash SIMT-threads executing different instructions (e.g., resulting from different if\slash then branches).
It is noted that generally speaking, shallow-vectorizations may also experience SIMD-waste\slash SIMT thread-divergence, e.g., in chemical kinetic integration due to varying internal solver time-step sizes~\cite{CurtisGPU:2017}; however, in this work shallow-vectorizations are largely unaffected by this concern \add{as the only major code-paths that differ between vector-lanes are high\slash low-temperature polynomial evaluations and differing pressures for P-Log reactions, a relatively small concern as compared to differing internal ODE integration time-steps~\cite{CurtisGPU:2017}}.

@@

\subsection{Thermo-chemical Source-Terms and Jacobian}
This new version of \texttt{pyJac} is capable of evaluating the thermo-chemical source-terms for using both the constant-pressure (\conp/) or constant-volume (\conv/) assumption\footnote{Note: in this context, the ``constant-pressure'' and ``constant-volume'' assumptions refer to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure or constant-volume reactive-flow simulation.}
In this section, we will only outline a brief summary of the system evaluated by \texttt{pyJac}; the interested reader is referred to the supplemental material for the complete derivations.

The thermo-chemical state vector derivation consists of the temperature, a non-constant thermodynamic state parameter (volume or pressure for \conp/ and \conv/, respectively), and the number of moles of all species except the last species in the chemical model which is typically taken to be the bath-gas (e.g., \ce{N2}):
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure, respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns/$ total species).

@@

The evolution of the thermo-chemical state vector is described by a set of chemical kinetic ordinary differential equations:
\loadeq{dstate}

@@

As in our previous work~\cite{Niemeyer:2016aa}, the correctness of the Jacobian was determined by comparison to the Jacobian obtained by automatic differentiation of the source-terms kernel generated by \texttt{pyJac} via the \texttt{Adept} software library~\cite{adept-v11,hogan2014fast}.
The reasons for this choice are explained fully in~\cite{Niemeyer:2016aa}, but broadly speaking automatic differentiation provides relatively fast, highly accurate Jacobian matrix evaluation with minimal additional programming effort.
The discrepancy between the analytical and automatic differentiated Jacobians for thermo-chemical state $k$, denoted by $\mathcal{J}_k$ and $\hat{\mathcal{J}}_k$ respectively, is determined by the relative error Frobenius norm over all Jacobian indices $i, j$:
\begin{equation}
 \label{e:jac_error_base}
 E_{\text{rel}, k} = \left\lVert \frac{\hat{\mathcal{J}}_{ijk} - \mathcal{J}_{ijk}}{\hat{\mathcal{J}}_{ijk}} \right\rVert_{i,j,F}
\end{equation}

To avoid large relative errors of very small nonzero Jacobian elements due to accumulation of floating point error, the Frobenius norm of the automatically differentiated Jacobian is calculated over all thermo-chemical states $k$:
\begin{equation}
 \label{e:thresh}
 \mathcal{T} = \left\lVert \mathcal{\hat{J}} \right\rVert_{k, F}
\end{equation}
The error statistics reported in this section are then based only on matrix elements where $\mathcal{J}_{ijk} \ge \frac{\mathcal{T}}{\mathcal{C}}$, where $\mathcal{C}$ is the threshold parameter; this filtered form of~\cref{e:jac_error_base} is denoted $E_{\mathcal{C},k}$.
Finally, the Frobenius norm is calculated over all the states in the PaSR thermo-chemical condition database:
\begin{equation}
 \label{e:thresholded_error}
 E_{\mathcal{C}} = \left\lVert E_{\mathcal{C},k} \right\rVert_{k, F}
\end{equation}

>>>>

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform, \texttt{pyJac}, for single-instruction, multiple-data (SIMD) and single-instruction, multiple thread (SIMT) vectorized \revise{constant-pressure\slash volume} thermochemical source-term and sparse\slash dense chemical kinetic Jacobian evaluation has been developed.% Can you introduce this a bit more? good to start with at least one sentence of introduction. also, this is passive and quite long\slash complex
Selected chemical kinetic models covering the range of sizes typically used in reactive-flow simulations were used for demonstration.
A new formulation of the chemical kinetic governing equations was derived and validated, \revise{resulting in} Jacobian sparsities of \SIrange{28.6}{92.0}{$\percent$} for the tested models.
Speedups of \SIrange{3.40}{4.08}{$\times$} were found for shallow-vectorized OpenCL source-rate evaluation \revise{compared with} a parallel OpenMP code on an \avx/ central processing unit (CPU), increasing to \SIrange{6.63}{9.44}{$\times$} and \SIrange{3.03}{4.23}{$\times$} for sparse and dense chemical kinetic Jacobian evaluation, respectively.
In addition, the performance of a Nvidia \gpunew/ graphics processing unit (GPU) was compared to a Nvidia \gpuold/ GPU finding speedups of \SIrange{1.40}{1.88}{$\times$}, \SIrange{1.10}{1.59}{$\times$}, and \SIrange{1.36}{3.0}{$\times$} for source-rate, sparse, and dense Jacobian evaluation, respectively. %is this sentence necessary in the abstract?
Furthermore, the effect of data-ordering was investigated and \revise{a} storage pattern specifically formulated for vectorized evaluation was proposed; as well, the effect of the \revise{constant pressure\slash volume} assumptions and varying vector widths were studied on source-term evaluation performance.
Speedups over a first-order finite-difference approach reached \SIrange{17.60}{45.13}{$\times$} and \SIrange{55.11}{245.63}{$\times$} for sparse\slash dense evaluation on the CPU \slash GPU respectively,%not sure what numbers refer to- sparse, dense, CPU, GPU?
while dense Jacobian evaluation was up to \SI{19.56}{$\times$} and \SI{2.84}{$\times$} times faster than a previous version of \texttt{pyJac} on a CPU and GPU, respectively.
Finally, future directions for vectorized chemical kinetic evaluation and sparse linear-algebra techniques were discussed.
\end{abstract}

@@

A host of techniques have been developed to lessen the computational demand of chemical kinetic calculations while maintaining fidelity, \revise{falling} broadly into \revise{three} categories: removal of unimportant species and reactions~\cite{Lu:2006bb,Pepiot-Desjardins:2008,Hiremath:2010jw,Niemeyer:2010bt,Curtis:2015},
lumping of species with similar thermochemical properties~\cite{Lu:2007,Ahmed:2007fa,Pepiot:2008kq},
and time-scale methods that reduce numerical stiffness~\cite{Maas:1992ws,Lam:1994ws,Lu:2001ve,Gou:2010}.
\revise{We refer interested readers} to the recent review by Tur\'anyi and Tomlin~\cite{turanyi2016analysis} for a comprehensive overview.

@@

Recognizing the need to accelerate chemical-kinetic Jacobian evaluation and factorization, a number of recent works have been published on \revise{constructing analytical Jacobian matrices}; although as will be discussed at the end of this section, \revise{here we offer} several key improvements over past efforts.
\add{Schwer et al.~\cite{SCHWER2002270} were among the first to recognize the critical importance of a sparse analytical Jacobian to accelerate chemical kinetic simulations.}
\revise{Later, Safta et al.~\cite{Safta:2011vn} developed the \texttt{TChem} software package}, which was one of the first developed that provides analytical Jacobian evaluation. However, \texttt{TChem} has several limitations, including incompatibility with modern reaction types---i.e., pressure dependent Arrhenius (or P-Log) and Chebyshev reactions---and its lack of thread-safety to enable parallel execution~\cite{Curtis2017:tchem}.
Youssefi~\cite{Youssefi:2011tm} explored the importance of analytical Jacobian matrices for time-scale analysis techniques as well as their effect on computational efficiency in zero-dimensional homogeneous reactor simulations.
Bisetti~\cite{Bisetti:2012jw} developed an isothermal, isobaric analytical Jacobian code-generation utility;
\revise{this approach significantly increases Jacobian sparsity, although the chosen isothermal assumption is not typical in most combustion simulations.}
In the same work Bisetti \revise{also} provided a novel way to compute dense matrix-vector multiplications resulting from a change of system variables without \revise{storing} the full Jacobian.
Perini et al.~\cite{Perini:2012gy} developed an analytical Jacobian code for constant-volume combustion, with additional options to increase sparsity (at the expense of strict correctness) and \revise{tabulate} temperature-dependent properties; \revise{they reported} an \SI{80}{\percent} speedup over a finite-difference-based Jacobian when used in a multidimensional reactive-flow simulation.
Gao et al.~\cite{GAO2015287} derived a sparse analytical Jacobian, but did not validate it outside the context of use with an implicit-integration technique.
In addition, since the Jacobian was based on an over-constrained system~\cite{HANSEN2018257}, the effect on strict conservation of mass\slash energy was not studied.

\revise{Recently, some groups have developed frameworks for constructing analytical Jacobians for evaluation on modern SIMD or SIMT processors.}
Dijkmans et al.~\cite{Dijkmans:2014bb} developed a GPU-based analytical Jacobian code with optional tabulation of temperature-dependent properties, and \revise{showed} speedups up to \SI{120}{$\times$} for zero-dimensional chemical kinetic integration with large chemical models (\textasciitilde\num{3000} species).
\add{Bauer et al.~\cite{Bauer:2014} used warp-specialization to improve GPU-vectorization over a standard data-parallel vectorization approach; they achieved speedups of up to \SIrange{2.81}{3.75}{$\times$}, \SIrange{1.91}{2.58}{$\times$}, and \SIrange{1.4}{1.5}{$\times$} for evaluating viscosity, species diffusion, and chemical source terms, respectively}.
Niemeyer et al.~\cite{Niemeyer:2016aa} created and validated the open-source analytical chemical kinetic Jacobian code-generator, \texttt{pyJac}, which supports parallel execution on CPUs and SIMT execution on GPUs; \revise{\texttt{pyJac} enables} a speedup of \SIrange{3}{7.5}{$\times$} over a finite-difference Jacobian on the CPU.

\add{Relevant to all of the aforementioned efforts, Hansen and Sutherland~\cite{HANSEN2018257} explored the choice of thermochemical state vectors and the resulting effect on consistency and errors in conserved properties such as mass and energy.
They also characterized how the choice of state vector affects implicit\slash linearly implicit integration algorithms and chemical mode analysis techniques.
Overall they found that while many literature Jacobian formulations are not strictly correct or over-specified, such flaws negligibly affect Newton--Krylov methods---perhaps because the incorrect Jacobian reasonably approximates the true Jacobian.
On the other hand, linearly implicit algorithms like Rosenbrock methods and analysis techniques like chemical explosive mode analysis~\cite{lu_yoo_chen_law_2010} need accurate and correct Jacobians.
}

A number of recent works have investigated \revise{using} high-performance SIMT devices \revise{like GPUs} to accelerate reactive-flow and chemical kinetics simulations.
Spafford et al.~\cite{Spafford:2010aa} coupled GPU-based chemical source-term evaluation with an explicit direct numerical simulation code, achieving an order of magnitude speedup \delete{on a Tesla C1060} compared to a \add{CPU-based} serial implementation \delete{on a AMD-Opteron processor}.
Shi et al.~\cite{Shi:2011aa} combined GPU-based chemical kinetic source-term evaluation and Jacobian factorization with two implicit CPU solvers, achieving an order-of-magnitude speedup for homogeneous reactor simulations of large chemical models\delete{on a Tesla 2050 GPU} over a serial\delete{Intel i7 930} CPU implementation.
Niemeyer et al.~\cite{Niemeyer:2011aa} implemented an explicit solver for non-stiff chemistry on a Tesla C2075 GPU, achieving a speedup of nearly two orders of magnitude over a sequential \add{CPU} code\delete{on an Intel Xeon X5650 CPU}.
Shi et al.~\cite{Shi:2012aa} proposed a strategy for chemical-kinetic integration in three-dimensional reactive-flow simulations, where a traditional implicit integrator handled the stiffest computational cells on a CPU and a stabilized-explicit solver \delete{on a Tesla C2050 GPU} solved the less-stiff cells on a GPU; this hybrid solution technique \revise{performs} \SIrange{11}{46}{$\times$} faster than the implicit CPU solver alone for simulation of a premixed diesel engine.
Le et al.~\cite{Le2013596} found a \SIrange{30}{50}{$\times$} speedup for a GPU-based shock-capturing reactive-flow code\delete{ on a Tesla C2070} as compared with a sequential \add{CPU} version of the same\delete{on an Intel Xeon X5650 CPU}.
Stone and Davis~\cite{Stone:2013aa} investigated a GPU-based version of a common implicit integrator (VODE~\cite{Brown:1989vl}), finding an order-of-magnitude speedup over a serial CPU implementation.
Niemeyer and Sung~\cite{Niemeyer:2014aa} developed a GPU-based stabilized explicit integrator for use with moderately-stiff chemical kinetics, achieving an order of magnitude speedup\delete{on a Tesla C2075 GPU} over a multithreaded VODE solver on a six-core\delete{Intel X5650} CPU.
Sewerin and Rigopoulos~\cite{Sewerin20151375} studied a fifth-order implicit Runge--Kutta solver on both consumer-grade and high-end GPUs\slash CPUs; the high-end GPU solver was at best \SI{1.8}{$\times$} slower than the high-end CPU version running on \num{16} cores.
Yonkee and Sutherland~\cite{Yonkee2016} implemented GPU-accelerated evaluations of thermodynamic parameters, multicomponent transport properties, and species production rates for solving partial differential equations (PDEs) \delete{on an Intel Xeon E5-2680} \add{both the CPU} and \delete{a Tesla K20} GPU.
In evaluating the thermochemical properties, speedups between \SIrange{8}{13}{$\times$} were found (over a serial version of the same code) while running on 16 CPU cores, and \SIrange{20}{40}{$\times$} on the GPU; speedups of \textasciitilde\SI{9}{$\times$} and \textasciitilde\SI{25}{$\times$} were found in solution of a partially premixed methanol flame PDE on 16 CPU cores and the GPU, respectively.\todo{revise this sentence, fairly awkward}{}
Curtis et al.~\cite{CurtisGPU:2017} implemented a fifth-order implicit Runge--Kutta method~\cite{wanner1991solving}, as well as two fourth-order exponential integration techniques~\cite{Hochbruck:1998,Hockbruck:2009} paired with an analytical Jacobian code~\cite{Niemeyer:2016aa}\delete{on a Tesla C2075} \add{on the GPU and CPU}.
The \add{GPU-based} implicit Runge--Kutta method was roughly equivalent to a standard implicit integrator~\cite{Hindmarsh:2005} running on \numrange{12}{38}\delete{Intel Xeon E5-4640 v2} CPU cores for two relatively small chemical models \delete{with an integration time step of $10^{-6}$ s}.

@@

When storing arrays for a chemical kinetic model, the data-storage layout and vectorization patterns are critical to achieving high-performance code.
\Cref{F:mem} depicts an example data array with $K$ rows and $N$ columns, and index ($i$, $j$) corresponding to the $i$th row and $j$th column.
For example, the concentration of species $j$ for the $i$th thermochemical state would be stored in $[C]_{i, j}$ with $1 \le i \le N_{state}$ (the number of thermochemical states considered for evaluation), and $1 \le j \le \ns/$ (the number of species in the model).
The stored concentrations would then have $K = N_{\text{state}}$ rows, and $N = N_{\text{sp}}$ columns.

In the ``C'' (C row-major) format, the concentrations of all species for a single thermochemical condition $i$ would be stored sequentially in memory, i.e., with  $[C]_{1, 1}$ in index \num{1}, $[C]_{1, 2}$ in index \num{2}, etc., as shown in \cref{F:row_major}.
Conversely in the ``F'' (Fortran column-major) format, the concentrations of a single species $j$ over all thermochemical states are adjacent in memory, corresponding to storing $[C]_{1, 1}$ in index \num{1}, $[C]_{2, 1}$ in index \num{2} and so on, as shown in \cref{F:column_major}.
This ordering---along with the device (CPU, GPU, etc.\@) and vectorization pattern in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD\slash SIMT vectorization (also referred to as ``per-thread'' in previous works~\cite{Stone:2013aa} using GPUs), each SIMD-lane (or SIMT-thread) in a compute-unit evaluates the thermochemical source-terms\slash chemical kinetic Jacobian for a different thermochemical state.
If the data is stored in ``F''-order, the SIMD-lanes\slash SIMT-thread accessing $[C]_{1, j}\ldots[C]_{N_v, j}$---the concentration of species $j$ for states $1, 2,\ldots N_v$, where $N_v$ is the SIMD vector-width or the number of threads in a SIMT warp---will load sequential locations in memory.
The first $(j+1)$th species concentration, $[C]_{1, j+1}$, will be $N_{\text{state}}$ memory locations away; this increases the likelihood of cache-misses on the CPU~\cite{gray2000rules}, but conversely is well suited to coalesced memory accesses on the GPU~\cite{Nvidia:2018}.

In a \textit{deep}-SIMD\slash SIMT vectorization (also referred to as ``per-block'' in previous GPU works (e.g.,~\cite{Stone:2013aa,CurtisGPU:2017}), a compute-unit utilizes its SIMD-lanes\slash SIMT-threads cooperatively to evaluate the thermochemical source-terms for a single thermochemical state; thus SIMD-lanes loading $[C]_{1, j} \ldots [C]_{1, j + N_v}$---the species concentrations for state 1, species $j \ldots j + N_v$---will access sequential memory locations if the data is stored in ``C''-order.
Further, in ``C''-ordering the maximum distance between any two species concentrations within the same thermochemical state is $N_{\text{sp}}$, with $N_{\text{sp}} \ll N_{\text{state}}$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU, but may lead to uncoalesced memory-accesses on the GPU.
In addition, a deep vectorization requires both synchronization between SIMD-lanes\slash SIMT-threads via memory-fences\slash barriers (an expensive operation), and may result in SIMD-waste\slash SIMT thread-divergence, caused by different SIMD-lanes\slash SIMT-threads executing different instructions (e.g., resulting from different if\slash then branches).
It is noted that generally speaking, shallow-vectorizations may also experience SIMD-waste\slash SIMT thread-divergence, e.g., in chemical kinetic integration due to varying internal solver time-step sizes~\cite{CurtisGPU:2017}; however, in this work shallow-vectorizations are largely unaffected by this concern \add{as the only major code-paths that differ between vector-lanes are high\slash low-temperature polynomial evaluations and differing pressures for P-Log reactions, a relatively small concern as compared to differing internal ODE integration time-steps~\cite{CurtisGPU:2017}}.

@@

\subsection{thermochemical Source-Terms and Jacobian}
This new version of \texttt{pyJac} is capable of evaluating the thermochemical source-terms for using both the constant-pressure (\conp/) or constant-volume (\conv/) assumption\footnote{Note: in this context, the ``constant-pressure'' and ``constant-volume'' assumptions refer to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure or constant-volume reactive-flow simulation.}
In this section, we will only outline a brief summary of the system evaluated by \texttt{pyJac}; the interested reader is referred to the supplemental material for the complete derivations.

The thermochemical state vector derivation consists of the temperature, a non-constant thermodynamic state parameter (volume or pressure for \conp/ and \conv/, respectively), and the number of moles of all species except the last species in the chemical model which is typically taken to be the bath-gas (e.g., \ce{N2}):
\loadeq{state}
where $T$ is the temperature, $V$ and $P$ the volume and pressure, respectively, and $n_j$ the number of moles of the $j$th species in the model (containing $\ns/$ total species).

@@

The evolution of the thermochemical state vector is described by a set of chemical kinetic ordinary differential equations:
\loadeq{dstate}

@@

As in our previous work~\cite{Niemeyer:2016aa}, the correctness of the Jacobian was determined by comparison to the Jacobian obtained by automatic differentiation of the source-terms kernel generated by \texttt{pyJac} via the \texttt{Adept} software library~\cite{adept-v11,hogan2014fast}.
The reasons for this choice are explained fully in~\cite{Niemeyer:2016aa}, but broadly speaking automatic differentiation provides relatively fast, highly accurate Jacobian matrix evaluation with minimal additional programming effort.
The discrepancy between the analytical and automatic differentiated Jacobians for thermochemical state $k$, denoted by $\mathcal{J}_k$ and $\hat{\mathcal{J}}_k$ respectively, is determined by the relative error Frobenius norm over all Jacobian indices $i, j$:
\begin{equation}
 \label{e:jac_error_base}
 E_{\text{rel}, k} = \left\lVert \frac{\hat{\mathcal{J}}_{ijk} - \mathcal{J}_{ijk}}{\hat{\mathcal{J}}_{ijk}} \right\rVert_{i,j,F}
\end{equation}

To avoid large relative errors of very small nonzero Jacobian elements due to accumulation of floating point error, the Frobenius norm of the automatically differentiated Jacobian is calculated over all thermochemical states $k$:
\begin{equation}
 \label{e:thresh}
 \mathcal{T} = \left\lVert \mathcal{\hat{J}} \right\rVert_{k, F}
\end{equation}
The error statistics reported in this section are then based only on matrix elements where $\mathcal{J}_{ijk} \ge \frac{\mathcal{T}}{\mathcal{C}}$, where $\mathcal{C}$ is the threshold parameter; this filtered form of~\cref{e:jac_error_base} is denoted $E_{\mathcal{C},k}$.
Finally, the Frobenius norm is calculated over all the states in the PaSR thermochemical condition database:
\begin{equation}
 \label{e:thresholded_error}
 E_{\mathcal{C}} = \left\lVert E_{\mathcal{C},k} \right\rVert_{k, F}
\end{equation}
